---
title: "Final Presentation"
subtitle: "Apples and pears? Impacts of using the Jacobian scaled conditional Akaike Information Criteria for model selection in linear mixed models. Advanced Small Area Estimation (Prof. Schmid)"
author: "Niklas Ippisch"
date: "07.22.2026"
format: beamer
#incremental: TRUE
---

```{r}
#| echo: FALSE
#| warning: FALSE
library(tidyverse)
library(ggpubr)
library(kableExtra)
```


## Motivation

- Situation: Linear Mixed Model with $y_i = X_i\beta + Z_iu_i + \epsilon_i$

- Adequate set of variables essential for good model fit

- Data driven transformation can improve the model fit

- Transformations especially important if normality assumption(s) $u_i \sim N(0, G)$ and/or $\epsilon_i \sim N(0, \sigma^2I_n)$ not fulfilled

- Problem: Variable selection and estimation of transformation parameter $\lambda$ affect each other:
  - Optimal set of covariates depends on transformation chosen and vice versa
  - Other papers suggest the usage of a conditional Akaike Information Criteria: $cAIC = -2log(g(y|\hat{\theta}, \hat{u})) + 2K_{MLE}$
  - Problem: Not scale invariant
  - Naïve solution: variable selection on original scale and transformation afterwards

## Proposed Solution I

- Idea: Change of variable approach: $l(y|\theta, u) = g(\tilde{y}|\theta, u) \cdot J(\lambda, y)$, with $l(y|\theta, u)$ being the conditional model density $\rightarrow$ calculation of K-L-divergence between conditional true and model density

- Solution: Scaling the $cAIC$ with the Jacobian $J$: $JcAIC = -2log(g(\tilde{y}|\hat{\theta}, \hat{u})) - 2log(J(\hat{\lambda}, y)) + 2BC$

- $\tilde{y}$ being the vector of transformed $y$ with transformation $T_\lambda(y)$

- $g$ being the density function of the approximating model

- $J$ being the Jacobian matrix: $J(\hat{\lambda}, y) = \prod^D_{i = 1} \prod^{N_i}_{j = 1} \frac{\partial \tilde{y}_{ij}}{\partial y_{ij}} = \prod^D_{i = 1} \prod^{N_i}_{j = 1} (y_{ij} + s)^{\lambda - 1}$

- $BC$ being a bias correction term, estimated by a bootstrap

## Proposed Solution II

Consider a Box-Cox transformation ($i = 1, ..., D \text{ and } j = 1, ..., N_i$):
$$
T_\lambda(y_{ij}) = \tilde{y} =
\begin{cases}
  \frac{(y_{ij} + s)^\lambda - 1}{\lambda}   & \text{if } \lambda \neq 0 \\
  log(y_{ij} + s)    & \text{if } \lambda = 0
\end{cases}
$$

1.  Start with full model with all $p$ variables and estimate $\hat{\lambda}$ for a Box-Cox transformation 

2.  For each step $s = 1, ..., S$:
    i)  Consider all possible models which exclude one predictor
    ii) Estimate $\hat{\lambda}$ for each candidate and calculate $JcAIC$ with estimated $\hat{\lambda}$
    iii) Compare $JcAIC$ values and choose model with lowest value as new optimal model
3.  Compare $JcAIC$ from 2.iii) with the previously optimal model from step $s-1$. Repeat until no better $JcAIC$ can be found.

## Methods

- Research question: Does the proposed simultaneous selection and estimation outperform the naive selection in different data scenarios?
- Approach I: Simulation Study
  - Four different data generation processes
  - Four different variable selection mechanisms
  - Inclusion of two irrelevant variables in the model
- Approach II: Application to real world data
  - Data from Mozambique
  - Estimation of head count ration and poverty gap based on naive approach vs. proposed solution
  
## Variable Selection Mechanisms

- "No": $T_\lambda(y_{ij}) = y$ with simple stepwise variable selection on original scale

- "Log": $T_\lambda(y_{ij}) = log(y)$ with simple stepwise variable selection on original scale

- "Box-Cox Naïve": First stepwise selection on original scale and Box-Cox transformation afterwards

- "Box-Cox Optimal": Stepwise variable selection using $JcAIC$ on Box-Cox transformed $y$


## Data generating processes

- $x_{1, ij} \sim N(\mu_i, 3^2) \text{ with } \mu_i \sim U[-3, 3], \newline x_{2, ij} \sim Bin(1, 0.8), x_{3, ij} \sim N(0, 1)$

- Additional irrelevant variables: $z_1 \sim N(1, 0.1^2)$ and $z_2 \sim Bin(1, 0.4)$

- Normal (Low): $400-10x_{1, ij} + 100x_{2, ij} - 10x_{3, ij} + u_i + e_{ij}$, $u_i \sim N(0, 30^2), e_{ij} \sim N(0, 60^2)$

- Normal (High): $400-10x_{1, ij} + 100x_{2, ij} - 10x_{3, ij} + u_i + e_{ij}$, $u_i \sim N(0, 10^2), e_{ij} \sim N(0, 20^2)$

- Log: $exp\{10-x_{1, ij} + x_{2, ij} - 0.5x_{3, ij} + u_i + e_{ij}\}$

- Box-Cox: $[(10-x_{1, ij} + x_{2, ij} - 0.5x_{3, ij} + u_i + e_{ij})\cdot(-0.5)+1]^{\frac{1}{-0.5}}$

## B = 50 vs. B = 200
```{=latex}
\begin{table}[]
\centering
\scriptsize
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrrrrr}
\toprule
dgp & transformation & mean\_JcAIC\_b50 & mean\_JcAIC\_b200 & mean\_lambda\_b50 & mean\_lambda\_b200 & proportion\_correct\_b50 & proportion\_correct\_b200 \\
\midrule
boxcox      & box.cox       & -2785.  & -2770.  & -0.483  & -0.483  & 0.8 & 0.9 \\
boxcox      & boxcox\_naive & -2667.  & -2638.  & -0.490  & -0.489  & 0.3 & 0.3 \\
boxcox      & log           & -2020.  & -2010.  & NaN     & NaN     & 1.0 & 1.0 \\
boxcox      & no            & 6470.   & 6480.   & NaN     & NaN     & 0.3 & 0.3 \\
log         & box.cox       & 14838.  & 14839.  & 0.00200 & 0.00197 & 0.6 & 0.8 \\
log         & boxcox\_naive & 14914.  & 14906.  & 0.00345 & 0.00336 & 0.2 & 0.2 \\
log         & log           & 14837.  & 14838.  & NaN     & NaN     & 0.6 & 0.8 \\
log         & no            & 21501.  & 21511.  & NaN     & NaN     & 0.2 & 0.2 \\
normal\_high & box.cox      & 7046.   & 7040.   & 1.26    & 1.35    & 0.6 & 0.6 \\
normal\_high & boxcox\_naive& 7098.   & 7065.   & 0.951   & 0.951   & 0.8 & 0.8 \\
normal\_high & log          & 7161.   & 7173.   & NaN     & NaN     & 0.8 & 0.8 \\
normal\_high & no           & 7126.   & 7137.   & NaN     & NaN     & 0.8 & 0.8 \\
normal\_low  & box.cox      & 8352.   & 8405.   & 0.963   & 0.939   & 0.3 & 0.3 \\
normal\_low  & boxcox\_naive& 8374.   & 8410.   & 0.939   & 0.939   & 0.7 & 0.7 \\
normal\_low  & log          & 8481.   & 8492.   & NaN     & NaN     & 0.8 & 0.7 \\
normal\_low  & no           & 8451.   & 8463.   & NaN     & NaN     & 0.7 & 0.7 \\
\bottomrule
\end{tabular}%
}
\caption{Comparison of results across B = 50 and  B = 200 replications for the Bootstrap of the BC-Term}
\end{table}
```


## Data generating processes

```{r}
#| echo: FALSE
load("data/dgps_R1.RData")
pop_normal_low <- sim_res_pop[["normal_low"]][[1]]
pop_normal_high <-sim_res_pop[["normal_high"]][[1]]
pop_log <- sim_res_pop[["log"]][[1]]
pop_boxcox <- sim_res_pop[["boxcox"]][[1]]

y_normal_low <- pop_normal_low |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_normal_low$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Normal (Low)",
       caption = "Blue line indicates median",
       x = "Target Variable",
       y = "")
y_normal_high <- pop_normal_high |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_normal_high$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Normal (High)",
       caption = "Blue line indicates median",
       x = "Target Variable",
       y = "")
y_log <- pop_log |> 
  filter(y < 50000) |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_log$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Log",
       caption = "Data tructuated due to visibility, y < 50,000; \n blue line indicates median",
       x = "Target Variable",
       y = "")
y_boxcox <- pop_boxcox |> 
  filter(y < 1) |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_boxcox$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Box-Cox",
       caption = "Data tructuated due to visibility, y < 1; \n blue line indicates median",
       x = "Target Variable",
       y = "")

ggarrange(y_normal_low, y_normal_high, y_log, y_boxcox, ncol = 2, nrow = 2)

```

## Simulation Setup

- $R = 300$ iterations

- $N = 10000$ observations in $D = 50$ domains, sampled via random sample ($n_d \sim Unif(0, 30)$)

- Population-generation and random sampling based on sample size in each iteration, but same for all methods

- Interval for estimating $\lambda$: $\hat{\lambda} \in \{-2, 2\}$

- Bias correction term estimated via bootstrap with $B = 50$ (paper: $B = 200$, slight differences as will be discussed in the term paper)

## Results: Optimal Lambda's

```{r}
#| echo: FALSE
#| warning: FALSE

load("data/simulation_study_results_processed.RData")
res_df <- res_df |> 
  mutate(transformation = case_when(
    transformation == "box.cox"       ~ "Box-Cox (Optimal)",
    transformation == "boxcox_naive"  ~ "Box-Cox (Naive)",
    transformation == "log"           ~ "Log",
    transformation == "no"            ~ "None",
    TRUE ~ transformation),
    dgp = case_when(
      dgp == "normal_low" ~ "Normal (Low)",
      dgp == "normal_high" ~ "Normal (High)",
      dgp == "log" ~ "Log",
      dgp == "boxcox" ~ "Box Cox")) |> 
  rename(DGP = dgp,
         Transformation = transformation)
res_df |> 
  filter(!is.na(opt_lambda)) |> 
  group_by(DGP, Transformation) |> 
  ggplot(aes(x = DGP, y = opt_lambda, fill = Transformation)) +
  geom_boxplot() +
  ylim(c(-1, 2.1)) +
  theme_minimal() +
  labs(x = "Data Generating Process",
       y = "Estimated Optimal Lambda",
       fill = "Variable Selection") +
  scale_fill_brewer(palette = "Set2")
  
```

## Results: Optimal Lambda's

- Expectation: $\hat{\lambda} \approx 1$ for Normal settings $\hat{\lambda} \approx 0$ for Log Setting, and $\hat{\lambda} \approx -0.5$ for Box-Cox-Setting

```{r}
t_lambda <- res_df |> 
  filter(!is.na(opt_lambda)) |> 
  group_by(DGP, Transformation) |> 
  summarise(Min = round(min(opt_lambda), 2),
            Mean = round(mean(opt_lambda), 2),
            Median = round(median(opt_lambda), 2),
            Max = round(max(opt_lambda), 2))
kable(t_lambda)
```


## Results: JcAIC

```{r}
res_df |> 
  filter(!is.na(opt_model)) |> 
  group_by(DGP, Transformation) |> 
  ggplot(aes(x = DGP, y = JcAIC, fill = Transformation)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(x = "Data Generating Process",
       y = "(J)cAIC",
       fill = "Variable Selection") +
  scale_fill_brewer(palette = "Set2")
```

## Results: JcAIC - zoomed into Normal Settings

```{r}
res_df |> 
  filter(!is.na(opt_model)) |> 
  filter(DGP == "Normal (High)" | DGP == "Normal (Low)") |> 
  group_by(DGP, Transformation) |> 
  ggplot(aes(x = DGP, y = JcAIC, fill = Transformation)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(x = "Data Generating Process",
       y = "(J)cAIC",
       fill = "Variable Selection") +
  scale_fill_brewer(palette = "Set2")
```


## Results: JcAIC

- Normal (Low):

```{r}
#| echo: FALSE
t_jcaic_nl <- res_df |> 
  filter(DGP == "Normal (Low)") |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_nl)
```

- Normal (High):

```{r}
#| echo: FALSE
t_jcaic_nh <- res_df |> 
  filter(DGP == "Normal (High)" & !is.na(opt_model)) |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_nh)
```

## Results: JcAIC

- Log:

```{r}
#| echo: FALSE
t_jcaic_log <- res_df |> 
  filter(DGP == "Log") |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_log)
```

- Box-Cox:

```{r}
#| echo: FALSE
t_jcaic_bc <- res_df |> 
  filter(DGP == "Box Cox") |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_bc)
```

## Results: Correctly identified variables

```{r}
res_df |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0, 1, 0)) |> 
  filter(!is.na(opt_model)) |> 
  group_by(DGP, Transformation) |> 
  summarise(mean = mean(correct_model)) |> 
  ggplot(aes(x = DGP, y = mean, fill = Transformation)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(x = "Data Generating Process",
       y = "Mean of correctly identified models",
       fill = "Transformation") +
  scale_fill_brewer(palette = "Set2")
```


## Results: Correctly identified variables

- Normal (Low)

```{r}
#| echo: FALSE

t_model_nl <- res_df |> 
  filter(DGP == "Normal (Low)") |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z1), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nl)

```

- Normal (High)

```{r}
#| echo: FALSE

t_model_nh <- res_df |> 
  filter(DGP == "Normal (High)" & !is.na(opt_model)) |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z1), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nh)

```

## Results: Correctly identified variables

- Log

```{r}
#| echo: FALSE

t_model_nl <- res_df |> 
  filter(DGP == "Log") |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z1), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nl)

```

- Box-Cox

```{r}
#| echo: FALSE

t_model_nh <- res_df |> 
  filter(DGP == "Box Cox") |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z1), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nh)

```

## Discussion

- More computational intensive variable selection using the $JcAIC$ mostly improves performance compared to naïve methods
    - Estimated $\hat{\lambda}$ close to true $\lambda$, but same/worse performance than naïve method
    - Same performance for Normal and Log setup and best performance for Box-Cox DGP
    - Moderate performance regarding variable selection for Normal settings, good for Log and Box-Cox
- Is it worth the trouble? Yes, if you have the computational power

$$
\begin{align}
\text{Thank you for your attention!} \nonumber
\end{align}
$$

## Literature

Krennmair, P., Würz, N., Schmid, T., & Tzavidis, N. (2026). Random forests and mixed effectsrandom forests for small area estimation of general parameters: A poverty mapping casestudy in mozambique. *Annals of Applied Statistics,20*(1), 809–832.https://doi.org/10.1214/25-AOAS2126

Lee, Y., Rojas-Perilla, N., Runge, M., & Schmid, T. (2023). Variable selection using conditionalAIC for linear mixed models with data-driven transformations. *Statistics and Computing,33*(1), 27.https://doi.org/10.1007/s11222-022-10198-9
