---
title: "Final Presentation"
subtitle: "Comparing apples and pears? Impacts of using the Jacobian scaled conditional Akaike Information Criteria for model selection in linear mixed models. Advanced Small Area Estimation (Prof. Dr. Schmid)"
author: "Niklas Ippisch"
date: "07.22.2026"
format: beamer
toc-depth: 3
toc-title: "Table of Contents"
toc: TRUE
#incremental: TRUE
---

```{r}
#| echo: FALSE
#| warning: FALSE
library(tidyverse)
library(ggpubr)
library(kableExtra)
```

# Motivation

## Motivation

- Situation: Linear Mixed Model with $y_i = X_i\beta + Z_iu_i + \epsilon_i$

- Adequate set of variables essential for good model fit

- Data driven transformation can improve the model fit

- Transformations especially important if normality assumption(s) $u_i \sim N(0, G)$ and/or $\epsilon_i \sim N(0, \sigma^2I_n)$ not fulfilled

- Problem: Variable selection and estimation of transformation parameter $\lambda$ affect each other:
  - Optimal set of covariates depends on transformation chosen and vice versa
  - Other papers suggest the usage of a conditional Akaike Information Criteria: $cAIC = -2log(g(y|\hat{\theta}, \hat{u})) + 2K_{MLE}$
  - Problem: Not scale invariant
  - Naive solution: variable selection on original scale and transformation afterwards
  
# Proposed solution

## Proposed Solution by Lee et al. (2023) I

- Idea: Change of variable approach: $l(y|\theta, u) = g(\tilde{y}|\theta, u) \cdot J(\lambda, y)$, with $l(y|\theta, u)$ being the conditional model density $\rightarrow$ calculation of K-L-divergence between conditional true and model density

- Solution: Scaling the $cAIC$ with the Jacobian $J$: $JcAIC = -2log(g(\tilde{y}|\hat{\theta}, \hat{u})) - 2log(J(\hat{\lambda}, y)) + 2BC$

- $\tilde{y}$ being the vector of transformed $y$ with transformation $T_\lambda(y)$

- $g$ being the density function of the approximating model

- $J$ being the Jacobian matrix: $J(\hat{\lambda}, y) = \prod^D_{i = 1} \prod^{N_i}_{j = 1} \frac{\partial \tilde{y}_{ij}}{\partial y_{ij}} = \prod^D_{i = 1} \prod^{N_i}_{j = 1} (y_{ij} + s)^{\lambda - 1}$

- $BC$ being a bias correction term, estimated by a bootstrap

## Proposed Solution by Lee et al. (2023) II

Consider a Box-Cox transformation ($i = 1, ..., D \text{ and } j = 1, ..., N_i$):
$$
T_\lambda(y_{ij}) = \tilde{y} =
\begin{cases}
  \frac{(y_{ij} + s)^\lambda - 1}{\lambda}   & \text{if } \lambda \neq 0 \\
  log(y_{ij} + s)    & \text{if } \lambda = 0
\end{cases}
$$

1.  Start with full model with all $p$ variables and estimate $\hat{\lambda}$ for a Box-Cox transformation 

2.  For each step $s = 1, ..., S$:
    i)  Consider all possible models which exclude one predictor
    ii) Estimate $\hat{\lambda}$ for each candidate and calculate $JcAIC$ with estimated $\hat{\lambda}$
    iii) Compare $JcAIC$ values and choose model with lowest value as new optimal model
3.  Compare $JcAIC$ from 2.iii) with the previously optimal model from step $s-1$. Repeat until no better $JcAIC$ can be found.

# Methods for the term paper

## Methods

- Research question: Does the proposed simultaneous selection and estimation outperform the naive selection in different data scenarios?
- Approach I: Simulation Study
  - Four different data generation processes
  - Four different variable selection mechanisms
  - Inclusion of two irrelevant variables in the model
- Hypotheses
  - "No" and "Log" transformation perform better in Normal and Log settings than the Box-Cox approaches
  - The proposed solution (Box-Cox (Optimal)) comes close to the performance of "No" and "Log"
  - Box-Cox (Optimal) outperforms the naive approach both in terms of $JcAIC$ and variable selection accuracy
  
## Variable Selection Mechanisms

- "No": $T_\lambda(y_{ij}) = y$ with simple stepwise variable selection on original scale

- "Log": $T_\lambda(y_{ij}) = log(y)$ with simple stepwise variable selection on original scale

- "Box-Cox Naive": First stepwise selection on original scale and Box-Cox transformation afterwards

- "Box-Cox Optimal": Stepwise variable selection using $JcAIC$ on Box-Cox transformed $y$


## Data generating processes

- $x_{1, ij} \sim N(\mu_i, 3^2) \text{ with } \mu_i \sim U[-3, 3], \newline x_{2, ij} \sim Bin(1, 0.8), x_{3, ij} \sim N(0, 1)$

- Additional irrelevant variables: $z_1 \sim N(1, 0.1^2)$ and $z_2 \sim Bin(1, 0.4)$

- Normal (Low): $y_{ij} = 400-10x_{1, ij} + 100x_{2, ij} - 10x_{3, ij} + u_i + e_{ij}$, $u_i \sim N(0, 30^2), e_{ij} \sim N(0, 60^2)$

- Normal (High): $y_{ij} = 400-10x_{1, ij} + 100x_{2, ij} - 10x_{3, ij} + u_i + e_{ij}$, $u_i \sim N(0, \boldsymbol{10}^2), e_{ij} \sim N(0, \boldsymbol{20}^2)$

- Log: $y_{ij} = exp\{10-x_{1, ij} + x_{2, ij} - 0.5x_{3, ij} + u_i + e_{ij}\}$

- Box-Cox: $y_{ij} = [(10-x_{1, ij} + x_{2, ij} - 0.5x_{3, ij} + u_i + e_{ij})\cdot(-0.5)+1]^{\frac{1}{-0.5}}$




## Data generating processes

```{r}
#| echo: FALSE
load("data/dgps_R1.RData")
pop_normal_low <- sim_res_pop[["normal_low"]][[1]]
pop_normal_high <-sim_res_pop[["normal_high"]][[1]]
pop_log <- sim_res_pop[["log"]][[1]]
pop_boxcox <- sim_res_pop[["boxcox"]][[1]]

y_normal_low <- pop_normal_low |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_normal_low$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Normal (Low)",
       caption = "Blue line indicates median",
       x = "Target Variable",
       y = "")
y_normal_high <- pop_normal_high |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_normal_high$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Normal (High)",
       caption = "Blue line indicates median",
       x = "Target Variable",
       y = "")
y_log <- pop_log |> 
  filter(y < 50000) |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_log$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Log",
       caption = "Data tructuated due to visibility, y < 50,000; \n blue line indicates median",
       x = "Target Variable",
       y = "")
y_boxcox <- pop_boxcox |> 
  filter(y < 1) |> 
  ggplot(aes(x = y)) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = median(pop_boxcox$y), color = "blue") +
  theme_minimal() +
  labs(subtitle = "Box-Cox",
       caption = "Data tructuated due to visibility, y < 1; \n blue line indicates median",
       x = "Target Variable",
       y = "")

ggarrange(y_normal_low, y_normal_high, y_log, y_boxcox, ncol = 2, nrow = 2)

```

## Simulation Setup

- $R = 300$ iterations

- $N = 10000$ observations in $D = 50$ domains, sampled via random sample ($n_d \sim Unif(0, 30)$)

- Population-generation and random sampling based on sample size in each iteration, but same for all methods

- Interval for estimating $\lambda$: $\hat{\lambda} \in \{-2, 2\}$

- Bias correction term estimated via bootstrap with $B = 50$ (paper: $B = 200$, slight differences as will be discussed in the term paper)

- 8 models with Log transformation did not converge in the Normal (High) setup

# Results

## Results: Optimal Lambda's

```{r}
#| echo: FALSE
#| warning: FALSE

load("data/simulation_study_results_processed.RData")
res_df <- res_df |> 
  mutate(transformation = case_when(
    transformation == "box.cox"       ~ "Box-Cox (Optimal)",
    transformation == "boxcox_naive"  ~ "Box-Cox (Naive)",
    transformation == "log"           ~ "Log",
    transformation == "no"            ~ "None",
    TRUE ~ transformation),
    dgp = case_when(
      dgp == "normal_low" ~ "Normal (Low)",
      dgp == "normal_high" ~ "Normal (High)",
      dgp == "log" ~ "Log",
      dgp == "boxcox" ~ "Box Cox"),
    jacobian = -clogLik - 0.5 * JcAIC + boot_BC) |> 
  rename(DGP = dgp,
         Transformation = transformation)
res_df |> 
  filter(!is.na(opt_lambda)) |> 
  group_by(DGP, Transformation) |> 
  ggplot(aes(x = DGP, y = opt_lambda, fill = Transformation)) +
  geom_boxplot() +
  ylim(c(-1, 2.1)) +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14)) +
  labs(x = "",
       y = "Estimated Optimal Lambda",
       fill = "Variable Selection") +
  scale_fill_brewer(palette = "Set2")
  
```

## Results: JcAIC

```{r}
res_df |> 
  filter(!is.na(opt_model)) |> 
  group_by(DGP, Transformation) |> 
  ggplot(aes(x = DGP, y = JcAIC, fill = Transformation)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14)) +
  labs(x = "",
       y = "(J)cAIC",
       fill = "Variable Selection") +
  scale_fill_brewer(palette = "Set2")
```

## Results: JcAIC - zoomed into Normal Settings

```{r}
res_df |> 
  filter(!is.na(opt_model)) |> 
  filter(DGP == "Normal (High)" | DGP == "Normal (Low)") |> 
  group_by(DGP, Transformation) |> 
  ggplot(aes(x = DGP, y = JcAIC, fill = Transformation)) +
  geom_boxplot() +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14)) +
  labs(x = "",
       y = "(J)cAIC",
       fill = "Variable Selection") +
  scale_fill_brewer(palette = "Set2")
```


## Correctly identified models

```{r}
#| echo: FALSE
res_agg <- res_df[0, ]
R <- 300
dgp_settings <- c("Normal (Low)", "Normal (High)", "Log", "Box Cox")

for(r in 1:R) {
  
  for(d in 1:length(dgp_settings)) {
    
    df_tmp <- res_df |> 
      filter(R == r & DGP == dgp_settings[d]) |> 
      arrange(JcAIC) |> 
      head(n = 1)
    res_agg <- rbind(res_agg, df_tmp)
  }
}

t_agg <- res_agg |> 
  group_by(DGP) |> 
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0),
         boxcoxopt = ifelse(Transformation == "Box-Cox (Optimal)", 1, 0),
         boxcoxnaive = ifelse(Transformation == "Box-Cox (Naive)", 1, 0),
         log = ifelse(Transformation == "Log", 1, 0),
         no = ifelse(Transformation == "None", 1, 0)) |> 
  summarise(No = round(mean(no), 2),
            Log = round(mean(log), 2),
            Naive = round(mean(boxcoxnaive), 2), 
            Optimal = round(mean(boxcoxopt), 2),
            Correct = round(mean(Correct), 2))

kable(t_agg, caption = "Percentage of models with lowest JcAIC per iteration and fraction of those which were correct.")

```

- Just half of them has identified the model correctly in the normal setups

- For Log and Box-Cox, high identification rates

- Box-Cox (Optimal) shows lowest $JcAIC$ in almost all cases of Box-Cox and half of the other settings

- Naive selection method does almost not yield lowest $JcAIC$

## Results: Correctly identified variables

```{r}
res_df |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0)) |> 
  filter(!is.na(opt_model)) |> 
  group_by(DGP, Transformation) |> 
  summarise(mean = mean(correct_model)) |> 
  ggplot(aes(x = DGP, y = mean, fill = Transformation)) +
  geom_bar(stat = "identity", position = "dodge") +
  ylim(0, 1) +
  theme_minimal() +
  theme(legend.position = "top",
        axis.text.x = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 14)) +
  labs(x = "",
       y = "Mean of correctly identified models",
       fill = "Transformation") +
  scale_fill_brewer(palette = "Set2")
```


## Zooming In: Normal (High) & Box-Cox (Optimal)

```{r}
#| echo: FALSE
t_normalhigh_boxcoxopt <- res_df |> 
  filter(DGP == "Normal (High)" & Transformation == "Box-Cox (Optimal)") |> 
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No")) |> 
  group_by(Correct) |> 
  summarise(Mean = round(mean(opt_lambda), 2),
            Median = round(median(opt_lambda), 2),
            SD  = round(sd(opt_lambda), 2), 
            x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z2), 2))
kable(t_normalhigh_boxcoxopt, caption = "Mean and Median Lambda as well as fraction of variables included per correctly identified model or not")
```
- $\hat{\lambda}$ of false models tends towards 2, for correct models towards 1
- But: $0.9 < \hat{\lambda} < 1.1$ does not ensure correct model specification

```{r}
t_lambda <- res_df |> 
  filter(DGP == "Normal (High)" & Transformation == "Box-Cox (Optimal)") |> 
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0),
         Lambda = ifelse(opt_lambda > 0.9 & opt_lambda < 1.1, 1, 0)) |> 
  group_by(Lambda) |> 
  summarise(Mean = round(mean(opt_lambda), 2),
            SD  = round(sd(opt_lambda), 2), 
            x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z2), 2),
            Correct = round(mean(Correct), 2))
kable(t_lambda, caption = "Mean and Median Lambda as well as fraction of variables included per lambda within [0.9, 1.1]")
```

## Closer Look I

```{r}
#| echo: FALSE

plot_jcaic_lambda1 <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         big_lambda = ifelse(opt_lambda > 1.7, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct), shape = big_lambda)) +
  geom_point(aes(y = JcAIC)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "JcAIC",
       color = "Model correctly identified?",
       shape = "Lambda > 1.7?") +
  scale_color_brewer(palette = "Set2")

plot_loglik_lambda1 <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         big_lambda = ifelse(opt_lambda > 1.7, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct), shape = big_lambda)) +
  geom_point(aes(y = clogLik)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "Log-Likelihood",
       y = "Log-Likelihood",
       color = "Model correctly identified?",
       shape = "Lambda > 1.7?") +
  scale_color_brewer(palette = "Set2")

plot_jacobi_lambda1 <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         big_lambda = ifelse(opt_lambda > 1.7, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct), shape = big_lambda)) +
  geom_point(aes(y = jacobian)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "Jacobian",
       y = "Jacobian",
       color = "Model correctly identified?",
       shape = "Lambda > 1.7?") +
  scale_color_brewer(palette = "Set2")

plot_BC_lambda1 <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         big_lambda = ifelse(opt_lambda > 1.7, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct), shape = big_lambda)) +
  geom_point(aes(y = boot_BC)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "Estimated Bias Correction Term",
       y = "Bias Correction Term",
       color = "Model correctly identified?",
       shape = "Lambda > 1.7?") +
  scale_color_brewer(palette = "Set2")

ggarrange(plot_jcaic_lambda1, plot_loglik_lambda1, plot_jacobi_lambda1, plot_BC_lambda1, common.legend = TRUE)
```

## Closer Look II

```{r}
#| echo: FALSE
t_correction <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(Model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Correct", "False"),
         Lambda = ifelse(opt_lambda < 0.90, "< 0.9", ifelse(opt_lambda > 1.1, "> 1.1", "Correct")),
         log_lik2 = (-2) * clogLik,
         corrections = (-2) * jacobian + 2 * boot_BC) |> 
  group_by(Model, Lambda) |> 
  summarise(Log_Lik = round(mean(log_lik2, na.rm = TRUE), 2), 
            Correction = round(mean(corrections, na.rm = TRUE), 2), 
            JcAIC = round(mean(JcAIC, na.rm = TRUE), 2)) |> 
  arrange(JcAIC)
kable(t_correction, caption = "Mean of Log-Likelihood, Correction Term, and JcAIC per correctly identified model and estimated lambda, ascending by JcAIC",
      col.names = c("Model", "Lambda", "-2*Log-Lik", "-2*J + 2*BC", "JcAIC")) 
```


- As reminder: $JcAIC = -2log(g(\tilde{y}|\hat{\theta}, \hat{u})) - 2log(J(\hat{\lambda}, y)) + 2BC$

- Correction Term might cancel out worse log-likelihood 

- Transformed models seemingly get preferred by the criterion

# Discussion

## Discussion

- Estimating $\hat{\lambda}$ worked fairly well for both the naive and optimal approach, except of the Normal (High) setup

- Box-Cox (Optimal) often yields models with lowest $JcAIC$

- Variable selection of Box-Cox (Optimal) works (very) well in Box-Cox and Log setting, but not in Normal settings

- Models with correct variables chosen have higher $JcAIC$ than wrong models with transformations

- Next steps: 
    - Further investigating path-dependency and skewed distributions
    - Simulating data again for Box-Cox (Optimal) with Normal settings and $B = 200$
    - Mathematical reasoning, esp. for wrong models with high likelihood

$$
\begin{align}
\text{Thank you for your attention!} \nonumber
\end{align}
$$

## Literature

Lee, Y., Rojas-Perilla, N., Runge, M., & Schmid, T. (2023). Variable selection using conditionalAIC for linear mixed models with data-driven transformations. *Statistics and Computing,33*(1), 27.https://doi.org/10.1007/s11222-022-10198-9

Reproduction materials: https://github.com/nippisch/Replication-Lee-et-al.

# Appendix

## Comparison B = 50 vs. B = 200 with R = 10
```{=latex}
\begin{table}[]
\centering
\scriptsize
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrrrrr}
\toprule
dgp & transformation & mean\_JcAIC\_b50 & mean\_JcAIC\_b200 & mean\_lambda\_b50 & mean\_lambda\_b200 & proportion\_correct\_b50 & proportion\_correct\_b200 \\
\midrule
boxcox      & box.cox       & -2785.  & -2770.  & -0.483  & -0.483  & 0.8 & 0.9 \\
boxcox      & boxcox\_naive & -2667.  & -2638.  & -0.490  & -0.489  & 0.3 & 0.3 \\
boxcox      & log           & -2020.  & -2010.  & NaN     & NaN     & 1.0 & 1.0 \\
boxcox      & no            & 6470.   & 6480.   & NaN     & NaN     & 0.3 & 0.3 \\
log         & box.cox       & 14838.  & 14839.  & 0.00200 & 0.00197 & 0.6 & 0.8 \\
log         & boxcox\_naive & 14914.  & 14906.  & 0.00345 & 0.00336 & 0.2 & 0.2 \\
log         & log           & 14837.  & 14838.  & NaN     & NaN     & 0.6 & 0.8 \\
log         & no            & 21501.  & 21511.  & NaN     & NaN     & 0.2 & 0.2 \\
normal\_high & box.cox      & 7046.   & 7040.   & 1.26    & 1.35    & 0.6 & 0.6 \\
normal\_high & boxcox\_naive& 7098.   & 7065.   & 0.951   & 0.951   & 0.8 & 0.8 \\
normal\_high & log          & 7161.   & 7173.   & NaN     & NaN     & 0.8 & 0.8 \\
normal\_high & no           & 7126.   & 7137.   & NaN     & NaN     & 0.8 & 0.8 \\
normal\_low  & box.cox      & 8352.   & 8405.   & 0.963   & 0.939   & 0.3 & 0.3 \\
normal\_low  & boxcox\_naive& 8374.   & 8410.   & 0.939   & 0.939   & 0.7 & 0.7 \\
normal\_low  & log          & 8481.   & 8492.   & NaN     & NaN     & 0.8 & 0.7 \\
normal\_low  & no           & 8451.   & 8463.   & NaN     & NaN     & 0.7 & 0.7 \\
\bottomrule
\end{tabular}%
}
\caption{Comparison of results across B = 50 and  B = 200 replications for the Bootstrap of the BC-Term}
\end{table}
```

## Results: Optimal Lambda's

- Expectation: $\hat{\lambda} \approx 1$ for Normal settings, $\hat{\lambda} \approx 0$ for Log Setting, and $\hat{\lambda} \approx -0.5$ for Box-Cox-Setting

```{r}
t_lambda <- res_df |> 
  filter(!is.na(opt_lambda)) |> 
  group_by(DGP, Transformation) |> 
  summarise(Min = round(min(opt_lambda), 2),
            Mean = round(mean(opt_lambda), 2),
            Median = round(median(opt_lambda), 2),
            Max = round(max(opt_lambda), 2))
kable(t_lambda, caption = "Summary statistics on estimated lambda per DGP and Transformation")
```


## Results: JcAIC

```{r}
#| echo: FALSE
t_jcaic_nl <- res_df |> 
  filter(DGP == "Normal (Low)") |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_nl, caption = "Summary statistics of JcAIC for Normal (Low)")
```


```{r}
#| echo: FALSE
t_jcaic_nh <- res_df |> 
  filter(DGP == "Normal (High)" & !is.na(opt_model)) |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_nh, caption = "Summary statistics of JcAIC for Normal (High)")
```

## Results: JcAIC

```{r}
#| echo: FALSE
t_jcaic_log <- res_df |> 
  filter(DGP == "Log") |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_log, caption = "Summary statistics of JcAIC for Log setting")
```

```{r}
#| echo: FALSE
t_jcaic_bc <- res_df |> 
  filter(DGP == "Box Cox") |> 
  group_by(Transformation) |> 
  summarise(Min = round(min(JcAIC), 2),
            Mean = round(mean(JcAIC), 2),
            Median = round(median(JcAIC), 2),
            Max = round(max(JcAIC), 2))
kable(t_jcaic_bc, caption = "Summary statistics of JcAIC for Box-Cox setting")
```

## Results: Correctly identified variables

```{r}
#| echo: FALSE

t_model_nl <- res_df |> 
  filter(DGP == "Normal (Low)") |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z2), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nl, caption = "Fraction of variables included in the models for Normal (Low) Setup")

```

```{r}
#| echo: FALSE

t_model_nh <- res_df |> 
  filter(DGP == "Normal (High)" & !is.na(opt_model)) |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z2), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nh, caption = "Fraction of variables included in the models for Normal (High) Setup")

```

## Results: Correctly identified variables

```{r}
#| echo: FALSE

t_model_nl <- res_df |> 
  filter(DGP == "Log") |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z2), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nl, caption = "Fraction of variables included in the models for Log Setup")

```

```{r}
#| echo: FALSE

t_model_nh <- res_df |> 
  filter(DGP == "Box Cox") |> 
  mutate(correct_model = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0)) |> 
  group_by(Transformation) |> 
  summarise(x1 = round(mean(x1), 2),
            x2 = round(mean(x2), 2),
            x3 = round(mean(x3), 2),
            z1 = round(mean(z1), 2),
            z2 = round(mean(z2), 2),
            Correct = round(mean(correct_model), 2))
kable(t_model_nh, caption = "Fraction of variables included in the models for Box-Cox Setup")

```

## Closer Look II

```{r}
#| echo: FALSE
#| eval: FALSE
comparison_all <- res_df |> 
  filter(!is.na(boot_BC) & !is.na(opt_model)) |> 
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0)) |> 
  group_by(DGP, Transformation) |> 
  summarise(lambda = mean(opt_lambda), JcAIC = mean(JcAIC), loglik = mean(clogLik), Jac = mean(jacobian),BC = mean(boot_BC))

comparison_correct <- res_df |> 
  filter(!is.na(boot_BC) & !is.na(opt_model)) |> 
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, 1, 0)) |> 
  group_by(DGP, Transformation, Correct) |> 
  summarise(lambda = mean(opt_lambda), JcAIC = mean(JcAIC), loglik = mean(clogLik), Jac = mean(jacobian),BC = mean(boot_BC))

plot_jcaic <- res_df |> 
  ggplot(aes(x = R, colour = Transformation)) +
  geom_line(aes(y = JcAIC)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "JcAIC")

plot_cloglik <- res_df |> 
  ggplot(aes(x = R, color = Transformation)) +
  geom_line(aes(y = clogLik)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "cLogLik")

plot_jacobian <- res_df |> 
  ggplot(aes(x = R, color = Transformation)) +
  geom_line(aes(y = jacobian)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "Jacobian")

plot_BC <- res_df |> 
  ggplot(aes(x = R, color = Transformation)) +
  geom_line(aes(y = boot_BC)) +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "BC")

ggarrange(plot_jcaic, plot_cloglik, plot_jacobian, plot_BC, common.legend = TRUE)
```

```{r}
#| echo: FALSE
means <- res_df |> 
  filter(Transformation == "None" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  summarise(jcaic = mean(JcAIC), loglik = mean(clogLik, na.rm = TRUE), jacobian = mean(jacobian, na.rm = TRUE), bc = mean(boot_BC, na.rm = TRUE))

plot_jcaic_lambda <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(real_lambda))) +
  geom_point(aes(y = JcAIC)) +
  geom_hline(yintercept = means[1, "jcaic"], color = "blue") +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "JcAIC",
       color = "Lambda within [0.9, 1.1]?") +
  scale_color_brewer(palette = "Set2")

plot_loglik_lambda <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(real_lambda))) +
  geom_point(aes(y = clogLik)) +
  geom_hline(yintercept = means[1, "loglik"], color = "blue") +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "Log-Likelihood",
       y = "Log-Likelihood",
       color = "Lambda within [0.9, 1.1]?")+
  scale_color_brewer(palette = "Set2")

plot_jacobi_lambda <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(real_lambda))) +
  geom_point(aes(y = jacobian)) +
  geom_hline(yintercept = means[1, "jacobian"], color = "blue") +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "Jacobian",
       y = "Jacobian",
       color = "Lambda within [0.9, 1.1]?")+
  scale_color_brewer(palette = "Set2")

plot_BC_lambda <- res_df |> 
  filter(Transformation == "Box-Cox (Optimal)" & (DGP == "Normal (High)" | DGP == "Normal (Low)")) |>
  mutate(real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(real_lambda))) +
  geom_point(aes(y = boot_BC)) +
  geom_hline(yintercept = means[1, "bc"], color = "blue") +
  facet_wrap(~DGP, scales = "free") +
  labs(title = "Estimated Bias Correction Term",
       y = "Bias Correction Term",
       color = "Lambda within [0.9, 1.1]?")+
  scale_color_brewer(palette = "Set2")

ggarrange(plot_jcaic_lambda, plot_loglik_lambda, plot_jacobi_lambda, plot_BC_lambda, common.legend = TRUE)
```

## Comparison with no transformation

```{r}
#| echo: FALSE

plot_jcaic_lambda2 <- res_df |> 
  filter((Transformation == "Box-Cox (Optimal)" | Transformation == "None") & DGP == "Normal (High)") |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct))) +
  geom_point(aes(y = JcAIC)) +
  facet_wrap(~Transformation, scales = "free") +
  labs(title = "JcAIC",
       color = "Model correctly identified?") +
  scale_color_brewer(palette = "Set2")

plot_loglik_lambda2 <- res_df |> 
  filter((Transformation == "Box-Cox (Optimal)" | Transformation == "None") & DGP == "Normal (High)") |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct))) +
  geom_point(aes(y = clogLik)) +
  facet_wrap(~Transformation, scales = "free") +
  labs(title = "Log-Likelihood",
       y = "Log-Likelihood",
       color = "Model correctly identified?") +
  scale_color_brewer(palette = "Set2")

plot_jacobi_lambda2 <- res_df |> 
  filter((Transformation == "Box-Cox (Optimal)" | Transformation == "None") & DGP == "Normal (High)") |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct))) +
  geom_point(aes(y = jacobian)) +
  facet_wrap(~Transformation, scales = "free") +
  labs(title = "Jacobian",
       y = "Jacobian",
       color = "Model correctly identified?") +
  scale_color_brewer(palette = "Set2")

plot_BC_lambda2 <- res_df |> 
  filter((Transformation == "Box-Cox (Optimal)" | Transformation == "None") & DGP == "Normal (High)") |>
  mutate(Correct = if_else(x1 == 1 & x2 == 1 & x3 == 1 & z1 == 0 & z2 == 0, "Yes", "No"),
         real_lambda = ifelse(opt_lambda > 0.90 & opt_lambda < 1.1, "Yes", "No")) |> 
  ggplot(aes(x = R, color = as.factor(Correct))) +
  geom_point(aes(y = boot_BC)) +
  facet_wrap(~Transformation, scales = "free") +
  labs(title = "Estimated Bias Correction Term",
       y = "Bias Correction Term",
       color = "Model correctly identified?") +
  scale_color_brewer(palette = "Set2")

ggarrange(plot_jcaic_lambda2, plot_loglik_lambda2, plot_jacobi_lambda2, plot_BC_lambda2, common.legend = TRUE)
```

## Not necessarily just skewed densities

```{r}
#| echo: FALSE
load("data/density_df.RData")
df_check <- res_df[res_df$opt_lambda > 1.7 & !is.na(res_df$opt_lambda), ]

r_lambda <- unique(df_check$R)

dens_df %>%
  mutate(lambda = if_else(sample %in% r_lambda, "Lambda > 1.7", "Lambda < 1.7")) |> 
  ggplot(aes(x = x, y = density, group = sample, color = lambda)) +
  geom_line(alpha = 0.3, linewidth = 0.5) +
  labs(
    x = "y",
    y = "Density",
    color = "High Lambda?") +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_color_brewer(palette = "Set2")
```