---
title: "Appendix I: Plausibility Checks for Data Generation"
author: "Niklas Ippisch"
format: html
---

```{r}
#| echo: FALSE
#| warning: FALSE
library(tidyverse)
library(ggpubr)
library(emdi)
```


This document contains several plausibility checks as appendix of the term paper in the course "Advanced Small Area Estimation". The goal of this appendix is to demonstrate the generation of a plausible data set, which will serve as base for the simulation studies conducted in the term paper. The document is inspired by the plausibility check documents supplemented for the paper by Rojas-Perilla et al. (2020).

In the following, data will be generated with the same code as used for the term paper. The only adjustment to the code used there is the saving of the population data, which is skipped in the simulation study.

```{r}
#| echo: FALSE
#| warning: FALSE

N <- 10000 # population size
D <- 50 # Number of domains
base_seed <- 123456 # base seed for reproducibility 
dgp_settings <- c("normal", "log", "boxcox") # data generating processes
sim_res_sample <- rep(list(vector("list", 1)), length(dgp_settings)) # results list
sim_res_pop <- rep(list(vector("list", 1)), length(dgp_settings)) # results list

# definition of help functions -------------------------------------------------

# sample function
sampler <- function(DAT, sample_size, D){
  
  # creation of empty dataframe
  smp <- DAT[0, ]
  
  for(d in 1:D){ # loop level 1: iterating over domains
    
    # subsetting all observations of domain d
    domain_rows <- DAT[DAT$domain == d, ]
    
    # saving the sample
    smp <- rbind(smp, domain_rows[sample(nrow(domain_rows), size = sample_size[d]), ], make.row.names = FALSE)
  }
  
  return(smp)
}

# data generation process function
dgp <- function(distribution, N, D) {
  
  # creating empty dataframe
  pop_r <- data.frame(id = 1:N)
  
  # randomly putting observations in one of the D domains
  pop_r[, "domain"] <- sample(1:D, size = N, replace = TRUE)
  
  # creating covariates constant across data settings
  pop_r[, "x3"] <- rnorm(n = N, mean = 0, sd = 1)
  pop_r[, "x4"] <- rbeta(n = N, shape1 = 2, shape2 = 2)
  pop_r[, "z1"] <- rnorm(n = N, mean = runif(n = 1, min = 95, max = 105), sd = 20)
  pop_r[, "z2"] <- rbinom(n = N, size = 1, prob = 0.4)
  
  if (distribution == "normal") { # if statement 1: normal distribution
    
    # creating the covariates, domain effects, and noise
    pop_r[, "x1"] <- rnorm(n = N, mean = runif(n = 1, min = -3, max = 3), sd = 3)
    pop_r[, "x2"] <- rbinom(n = N, size = 1, prob = 0.2)
    
    u <- rnorm(D, mean = 0, sd = 30)
    e <- rnorm(N, mean = 0, sd = 60)
    
    # creating dependent variable and binding it into data frame
    pop_r[, "y"] <- 1200 - 10 * pop_r$x1 + 100 * pop_r$x2 - 10 * pop_r$x3 + 0.03 * pop_r$x4 + u[pop_r$domain] + e
    
  } else if (distribution == "log") { # if statement 2: log distribution
    
    # creating the covariates, domain effects, and noise
    pop_r[, "x1"] <- rnorm(n = N, mean = runif(n = 1, min = 2, max = 3), sd = 2)
    pop_r[, "x2"] <- rbinom(n = N, size = 1, prob = 0.8)
    
    u <- rnorm(D, mean = 0, sd = 0.4)
    e <- rnorm(N, mean = 0, sd = 0.8)
    
    # creating dependent variable and binding it into data frame
    pop_r[, "y"] <- exp(10  - pop_r$x1 + pop_r$x2 - 0.5 * pop_r$x3 - 0.03 * pop_r$x4 + u[pop_r$domain] + e)
    
  } else if (distribution == "boxcox") { # if statement 3: boxcox
    
    # creating the covariates, domain effects, and noise
    pop_r[, "x1"] <- rnorm(n = N, mean = runif(n = 1, min = 2, max = 3), sd = 2)
    pop_r[, "x2"] <- rbinom(n = N, size = 1, prob = 0.8)
    
    u <- rnorm(D, mean = 0, sd = 0.4)
    e <- rnorm(N, mean = 0, sd = 0.8)
    
    # creating dependent variable and binding it into data frame
    pop_r[, "y"] <- ((10  - pop_r$x1 + pop_r$x2 - 0.5 * pop_r$x3 - 0.03 * pop_r$x4 + u[pop_r$domain] + e)*(-0.5) + 1)^(1/-0.5)
  }
  
  return(pop_r)
}

# simulation study -------------------------------------------------------------

r <- 1

for (j in 1:length(dgp_settings)) { # loop level 2: data generating processes
    
    # varying but reproducible seed
    set.seed(base_seed + 10 * r + 10 * j)
    
    # generating the population
    pop_r <- dgp(distribution = dgp_settings[j], N = N, D = D)
    
    # defining the sample size, drawing the sample, and saving it
    sample_size <- round(runif(n = D, min = 0, max = 30))
    sample_r <- sampler(DAT = pop_r, sample_size = sample_size, D = D)
    sim_res_sample[[j]] <- sample_r
    
    sim_res_pop[[j]] <- pop_r
    
  } # end loop level 2
  
pop_norm <- sim_res_pop[[1]]
pop_log <- sim_res_pop[[2]]
pop_boxcox <- sim_res_pop[[3]]

sample_norm <- sim_res_sample[[1]]
sample_log <- sim_res_sample[[2]]
sample_boxcox <- sim_res_sample[[3]]
```

# Density of dependent variable in population and sample

```{r}
#| echo: FALSE
#| warning: FALSE

dens_normal <- ggplot() +
  geom_density(data = sample_norm, aes(x = y, fill = "Sample"), alpha = 0.5) +
  geom_density(data = pop_norm, aes(x = y, fill = "Population"), alpha = 0.5) +
  labs(title = "Normal Setup",
       x = "Target variable",
       y = "Density",
       fill = "Data Source") +
  theme_minimal() +
  theme(legend.position = "top")

dens_log <- ggplot() +
  geom_density(data = sample_log, aes(x = y, fill = "Sample"), alpha = 0.5) +
  geom_density(data = pop_log, aes(x = y, fill = "Population"), alpha = 0.5) +
  labs(title = "Log Setup",
       x = "Target variable",
       y = "Density",
       fill = "Data Source") +
  xlim(0, 25000) +
  theme_minimal() +
  theme(legend.position = "top")

dens_boxcox <- ggplot() +
  geom_density(data = sample_boxcox, aes(x = y, fill = "Sample"), alpha = 0.5) +
  geom_density(data = pop_boxcox, aes(x = y, fill = "Population"), alpha = 0.5) +
  labs(title = "Box-Cox Setup",
       x = "Target variable",
       y = "Density",
       fill = "Data Source") +
  xlim(0, 2) +
  theme_minimal() +
  theme(legend.position = "top")

ggarrange(dens_normal, dens_log, dens_boxcox, nrow = 3, ncol = 1, common.legend = TRUE)


```


```{r}
#| echo: FALSE
#| warning: FALSE
data_ges <- bind_rows(pop_norm, pop_log, pop_boxcox,
                      sample_norm, sample_log, sample_boxcox, 
                      .id = "data_source")

data_ges |> 
  mutate(source = if_else(data_source < 4, "population", "sample")) |> 
  group_by(source, data_source) |> 
  summarise(mean = mean(y)) 
```

# Fitted EBP and Diagnostics

```{r}
#| echo: FALSE
#| warning: FALSE
ebp_norm <- ebp(y ~ x1 + x2, pop_data = pop_norm, pop_domains = "domain",
             smp_data = sample_norm, smp_domains = "domain", L = 50, 
             transformation = "box.cox")
summary(ebp_norm)
```

```{r}
#| echo: FALSE
#| warning: FALSE
plot(ebp_norm)
```



