---
title: 'Szenario tester - DGP Log-scale'
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

### Load required packages
library(emdi)
library(saeSim)
library(ggplot2)
### Step 1: Data generation with R package -------------------------------------------------

### a) Define input parameters and setup -------------------------------------------------
set.seed(100)
simruns <- 100  # Number of Monte-Carlo replications
Domains <- 50  # Number of Small Areas
# Define area-specific sample sizes
sample_size <- c(12, 23, 28, 14, 10, 23, 19, 25, 29, 10, 14, 18, 15, 20, 13, 12, 16, 
                 27, 20, 26, 27, 23, 12, 12, 11, 18, 17, 29, 11, 29, 17, 9, 14, 8, 
                 8, 18, 21, 21, 16, 16, 25, 13, 26, 19, 28, 20, 24, 9, 25, 21)
# Define area-specific population sizes
pop_size <- rep(200,Domains)
# First summary statistics
sum(pop_size) # Size of the population
sum(sample_size) # Size of the sample
summary(sample_size) # Summary of the area-specific sample sizes

# Help function (in order to keep the population in the data generation process)
calc_keepPop <- function(dat)
{
  attr(dat, "pop")=dat
  dat
}

### b) Define the data generation mechanism -------------------------------------------------
# Log-scale setup compare Rojas-Perilla et al. (2020) Table 6 
# Help functions for x, z, e and sampling
gen_XNorm=function(dat,m = dat$muD, s = 2){
  dat["x"]=rnorm(nrow(dat), mean = m, sd = s)
  return(dat)
}
gen_ZNorm=function(dat, m = dat$muD, s=1){
  dat["z"]=rnorm(nrow(dat), mean = 0, sd = s)
  return(dat)
}
gen_myE=function(dat, m = 0, s = sqrt(.8)){
  dat["e"] = rnorm(nrow(dat), mean = 0, sd = s)
  dat
}
sampler <- function(DAT){
  smp=as.data.frame(matrix(nrow=sum(sample_size) , ncol=ncol(DAT)))
  brd=append(0,cumsum(sample_size))
  for(i in 1:Domains){
    smp[((brd[i]+1):brd[i+1]),]=(DAT[DAT$idD==i,])[sample(1:sum(DAT$idD==i),size=sample_size[i]),]
  }
  attr(smp,"pop")=DAT
  colnames(smp)=colnames(DAT)
  return(smp)
}

# Define the components of one arbitrary population
setup=sim_base(data = base_id(nDomains=Domains,nUnits=pop_size)) %>%
  sim_gen(gen_generic(runif, min = 2, max = 3, groupVars="idD", name = "muD")) %>%
  sim_gen(gen_XNorm)       %>%
  sim_gen(gen_ZNorm)       %>% 
  as.data.frame %>%
  sim_gen(generator=gen_myE)         %>% 
  sim_gen_v(mean = 0, sd = 0.4)         %>% 
  sim_resp_eq(y = exp(10 - x - 0.5 * z + v + e))

# Generate "simruns" populations defined by the setup
Pop <- sim(setup,R=simruns)
# Generate "simruns" samples from the population with "sample_size" area-specific sample sizes
Pop <- lapply(Pop,sampler)
```

#### 1. Compare the density of the target variable in the population with that of a single Monte Carlo sample.
```{r echo = F, message = F, warning = F}
test_smp <- Pop[[1]]

# Extract the population data from the sample
test_pop <- attr(test_smp, "pop")

# Calculate the threshold, mean, and median
test_thold <- 0.6 * median(test_smp$y)
test_mean <- mean(test_smp$y)
test_median <- median(test_smp$y)


ggplot() +
  geom_density(data = test_smp, aes(x = y, fill = "Sample"), alpha = 0.5) +
  geom_density(data = test_pop, aes(x = y, fill = "Population"), alpha = 0.5) +
  geom_vline(xintercept = test_thold, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = test_mean, linetype = "dashed", color = "green", size = 1) +
  geom_vline(xintercept = test_median, linetype = "dashed", color = "blue", size = 1) +
  labs(title = "",
       x = "Target variable",
       y = "Density",
       fill = "Group") +
  scale_fill_manual(values = c("Sample" = "skyblue", "Population" = "orange")) +
  theme_minimal() +
  theme(legend.position = "top")
```

#### 2. Estimate an EBP model with box cox transformation and look at a summary.
```{r, echo = F, warning = F, message=F}
ebpBC <- ebp(y~x, pop_data = test_pop, pop_domains = "idD",
             smp_data = test_smp, smp_domains = "idD", L = 50, 
             transformation = "box.cox")
summary(ebpBC)
```

#### 3. Plot the EBP model and look at the residuals. 
```{r echo = F, warning = F, message=F}
plot(ebpBC)
```

#### 4. Take a closer look at some point estimates and check whether the results make sense. 
```{r echo = F, warning = F, message=F}
tableSum = rbind(summary(ebpBC$ind$Mean),
                 summary(ebpBC$ind$Head_Count),
                 summary(ebpBC$ind$Poverty_Gap),
                 summary(ebpBC$ind$Quintile_Share))
rownames(tableSum) = c("Mean","HCR", "PGAP", "QSR")
colnames(tableSum) = c("Min", "Q1", "Median", "Mean", "Q3", "Max")
#tableSum
library(knitr)
kable(tableSum, digits=4, caption ="Summary of indicators")
#ggarrange(mseMean_lin, mseMean, mseMean_nb5, mseMean_nb3, mseMean_nb1, common.legend = T, legend="bottom")

```

#### 5. Calculate the transformation parameter for the entire populations and examine the distribution.
```{r echo = F, warning = F, message=F}
all_ebps <- vector("list", length(Pop))
for(i in seq_along(Pop))
{
  tmp_smp <- Pop[[i]]
  tmp_pop <- attr(tmp_smp, "pop")  
  tmp_thold <- 0.6 * median(tmp_smp$y)
  
  all_ebps[[i]] <-ebp(y~x, pop_data = tmp_pop, pop_domains = "idD",
             smp_data = tmp_smp, smp_domains = "idD", L = 2, 
             transformation = "box.cox")
    
}
# Extract the optimal_lambda from each ebp result
all_lambdas <- sapply(all_ebps, function(list) list[["transform_param"]][["optimal_lambda"]])

# Combine the lambdas into a data frame
lambda_data <- data.frame(population = rep(1:length(all_ebps), each = length(all_lambdas[[1]])),
                          optimal_lambda = unlist(all_lambdas))

# Create a boxplot using ggplot2
ggplot(lambda_data, aes(y = optimal_lambda)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Optimal transformation parameter across populations",
       x = "",
       y = "Transformation parameter") +
  theme_minimal()

```

